---
title: "Test_Exponential Distribution"
author: "Jason Giacomo"
date: "November 7, 2015"
output: html_document
---

#Test - Exploring the Exponential Distribution

####Author: Jason Giacomo

##Overview:  
This analysis will explore how the distribution of sample means and variance
relate to the population mean and variance of the exponential distribution. 
Simulated random samplings from the exponential distribution will be taken and
the distribution of the means of these samplings will be discussed. The mean and
variance of the samplings will be compared to the population mean and variance.

<!--
    While the overview looks complete, it is rather bland. Maybe it needs to
    be in a more conversational tone?
-->

##Simulations:  
Simulated random samples will be taken from the exponential distribution using
the R function `rexp()`. The simulation will utilize a function
to evalulate the mean of 40 samples 1000 times. The rate of the exponential
distribution will be fixed at 0.2.

```{r simulation}
    expSim <- function(nSamples, nMeans){
        expMean <- NULL
        for(i in 1:nMeans) expMean <- c(expMean, mean(rexp(nSamples, 0.2)))
        expMean <- data.frame("means"=expMean)  # for ease of use with ggplot2
        return(expMean)
    }
```

This function is then used to generate 1000 means from 40 samples which is
plotted as a histogram to show the distribution of means.

```{r histogram}
    set.seed(333)
    myMeans <- expSim(40, 1000)
    str(myMeans)

    library(ggplot2)    
    histogram <- ggplot(myMeans, aes(x=means)) +
        geom_histogram(aes(y=..density..), binwidth=0.25, color="black",
                       fill="white") +
        geom_density(color="blue", fill="blue", alpha=0.2) +
        stat_function(fun=dnorm, color="red", args = list(mean = 1/0.2))
    
    print(histogram)
```



##Sample Mean vs. Theoretical Mean

##Sample Variance vs. Theoretical Variance

##Distribution of Sample Means